\chapter{Abstract}

This is an undergraduate project submitted in partial fulfillment of
the requirements for the degree of Systems Engineering at EAFIT
University.

The aim of this... %% at most 500 words

It is possible to conceptualize many concepts of functional
programming via the study of category theory. So this project aims to
develop a monographical study of category theory which allows to
describe and explain some of the applications to functional
programming.

The aim of this project is to describe and explain some of the
applications of category theory to functional programming. In
particular, we describe and explain the concepts of category theory
which allow us to conceptualize and better understand the following
concepts of functional programming: polymorphism, algebraic data
types, and the functor and monad type classes.

Our goal is to describe and explain some of the applications of
category theory to functional programming. More specifically, to
describe and explain the concepts of category theory which allow to
conceptualize and better understand the following concepts of
functional programming: polymorphism, algebraic data types, and the
functor and monad type classes.

%% Category --> Hask and Agda

%% Functor, endofunctor --> Functors

%% Initial algebras --> Algebraic data types

%% Natural transformations --> Parametrically polymorphic functions

\vspace{1em}
\noindent
Keywords: category, category theory, functional programming.

\clearemptydoublepage
