\chapter{Algebras}
\label{chap:algebras}

\epigraph{
  ``Curiouser and curiouser!''
}{---\textcite[23]{carroll-2004}}

In this chapter, which is based mainly on \parencite[ยง
2.1]{vene-2000}, we explore initial algebras over endofunctors and
their relation to algebraic data types in Haskell. In
Section \ref{sec:algebras} we describe algebras and initial algebras
over endofunctors, but restrict examples to the category \set. And in
Section \ref{sec:algebras-haskell} we consider algebraic data types in
Haskell from the point of view of initial algebras over endofunctors.

\section{Introduction}
\label{sec:algebras-introduction}

\todo{We are only concerned with inductive types, not coinductive.}

In functional programming, \texthaskell{foldr} is a standard function
that encapsulates common patterns of induction and recursion
concerning lists \parencite[355--356]{hutton-1999}. In Haskell,
\texthaskell{foldr} can be defined as follows\footnote{Actually, the
  type of the standard \texthaskell{foldr} function is
  \begin{equation*}
    \text{\texthaskell{foldr :: (a -> b -> b) -> b -> [a] -> b}}
    \text{.}
  \end{equation*}
}:
\begin{codehaskell}
foldr :: b -> (a -> b -> b) -> [a] -> b
foldr n c []     = n
foldr n c (x:xs) = c x (foldr n c xs)
\end{codehaskell}
That is, given a value \texthaskell{n} of type \texthaskell{b} and a
function \texthaskell{c} of type \texthaskell{a -> b -> b}, the
function \texthaskell{foldr n c} of type \texthaskell{[a] -> b}
replaces \texthaskell{[]} with \texthaskell{n} and \texthaskell{(:)}
with \texthaskell{c}. This amounts to saying that both
\texthaskell{[a]}, \texthaskell{[]}, and \texthaskell{(:)}, and
\texthaskell{b}, \texthaskell{n}, and \texthaskell{c} are algebras
over an endofunctor, and, more important, that the former constitute
the initial algebra over that endofunctor. As it turns out, this fact
uniquely determines both the type and the definition of the
\texthaskell{foldr} function.

This idea generalizes to algebraic data types in the sense that one
such type is the initial algebra over an endofunctor and that this
fact uniquely determines a function that encapsulates common patterns
of induction and recursion concerning the type.

\section{Algebras}
\label{sec:algebras}

We begin by describing algebras over endofunctors and then discuss some
examples which will be revisited later on.

\begin{definition}
  \label{def:algebra}

  %% \parencite[15]{vene-2000}

  Let $\func{F}: \cat{C} \to \cat{C}$ be an endofunctor in a category
  \cat{C}. An \func{F}-algebra $(a,\alpha)$ is an object
  \begin{equation}
    \label{eq:al-carrier}
    a \in \catO{C}
    \text{,}
  \end{equation}
  called the carrier of the algebra, and a morphism
  \begin{equation}
    \alpha: \funcO{F}(a) \to a \in \catM{C}
    \text{.}
  \end{equation}

\end{definition}

\begin{example}
  \label{ex:algebra-initial-object}

  %% \parencite[18]{vene-2000}

  Let \cat{C} be a category with an initial object $0 \in \catO{C}$.
  Then $(0,\idO{0})$ is an algebra over the identity functor (see
  Example \ref{ex:functor-identity}), that is, an \func{I}-algebra. In
  particular, in \set, $(\emptyset,\idO{\emptyset})$ is an
  \func{I}-algebra.

\end{example}

\begin{example}%%[Natural numbers in $\set$]
  \label{ex:algebra-natural-numbers}

  \parencite[18--19]{vene-2000}

  In \set, the natural numbers $\mathbb{N} = \{0,1,2,...\}$, along
  with the functions $\zero: 1 \to \mathbb{N}$ and $\suc: \mathbb{N}
  \to \mathbb{N}$, which can be joined to a function $[\zero,\suc]: 1
  + \mathbb{N} \to \mathbb{N}$, as illustrated by the diagram in
  Figure \ref{fig:coproduct-natural-numbers},
  \begin{figure}[htbp]
    \begin{center}
      \begin{tikzpicture}[node distance=4cm]
        \node (ab)               {$1 + \mathbb{N}$};

        \node (a)  [left of=ab]  {$1$};
        \node (b)  [right of=ab] {$\mathbb{N}$};
        \node (c)  [below of=ab] {$\mathbb{N}$};

        \draw [<-] (ab) to node [swap] {$\iota_{1}$} (a);
        \draw [<-] (ab) to node        {$\iota_{2}$} (b);

        \draw [<-] (c)  to node        {$\zero$} (a);
        \draw [<-] (c)  to node [swap] {$\suc$} (b);

        \draw [<-,dotted] (c)  to node [swap] {$[\zero,\suc]$} (ab);
      \end{tikzpicture}
    \end{center}
    \caption{}
    \label{fig:coproduct-natural-numbers}
  \end{figure}
  yield an algebra $(\mathbb{N},[\zero,\suc])$ over an endofunctor
  $\func{N}: \set \to \set$ whose object function $\funcO{N}:
  \catO\set \to \catO\set$ assigns to each set $A$ a set
  \begin{equation*}
    \funcO{N}(A) = 1 + A
    \text{,}
  \end{equation*}
  and whose morphism function $\funcM{N}: \catM\set \to \catM\set$
  assigns to each function $f: A \to B$ a function $\funcM{N}(f): 1 +
  A \to 1 + B$ such that
  \begin{equation*}
    \funcM{N}(f) \comp \iota_{1} = \iota_{1}
    \quad
    \text{and}
    \quad
    \funcM{N}(f) \comp \iota_{2} = \iota_{2} \comp f
    \text{,}
  \end{equation*}
  that is, such that
  \begin{equation*}
    \funcM{N}(f)(1,()) = (1,())
    \quad
    \text{and}
    \quad
    \funcM{N}(f)(2,x) = (2,f(x))
  \end{equation*}
  for all $x \in A$. If necessary, see Examples
  \ref{ex:initial-terminal-objects-set} and \ref{ex:coproduct-set}
  (that is, initial objects and coproducts in \set, respectively).

\end{example}

\begin{example}%%[Lists
  \label{ex:algebra-lists}

  %% \parencite[19--20]{vene-2000}

  In \set, lists can be represented as algebras over endofunctors. For
  a given set $A$,
  \begin{equation*}
    \nil: 1 \to \listof{A}
    \quad
    \text{and}
    \quad
    \cons: A \times \listof{A} \to \listof{A}
  \end{equation*}
  can be joined to a function $[\nil,\cons]: 1 + A \times \listof{A}
  \to \listof{A}$, as illustrated by the diagram in Figure
  \ref{fig:coproduct-lists}.
  \begin{figure}[htbp]
    \begin{center}
      \begin{tikzpicture}[node distance=4cm]
        \node (ab)               {$1 + A \times \listof{A}$};

        \node (a)  [left of=ab]  {$1$};
        \node (b)  [right of=ab] {$A \times \listof{A}$};
        \node (c)  [below of=ab] {$\listof{A}$};

        \draw [<-] (ab) to node [swap] {$\iota_{1}$} (a);
        \draw [<-] (ab) to node        {$\iota_{2}$} (b);

        \draw [<-] (c)  to node        {$\nil$} (a);
        \draw [<-] (c)  to node [swap] {$\cons$} (b);

        \draw [<-,dotted] (c)  to node [swap] {$[\nil,\cons]$} (ab);
      \end{tikzpicture}
    \end{center}
    \caption{}
    \label{fig:coproduct-lists}
  \end{figure}
  In this way, $(\listof{A},[\nil,\cons])$ is an algebra over an
  endofunctor $\func{L}^{A}: \set \to \set$ whose object function
  $\funcO{L}^{A}: \catO{\set} \to \catO{\set}$ assigns to each set $B$
  a set
  \begin{equation}
    \funcO{L}^{A}(B) = 1 + A \times B
    \text{,}
  \end{equation}
  and whose morphism function $\funcM{L}^{A}: \catM{\set} \to
  \catM{\set}$ assigns to each function $g: B \to C$ a function
  $\funcM{L}^{A}(g): 1 + A \times B \to 1 + A \times C$ such that
  \begin{equation}
    \funcM{L}^{A}(g)(1,()) = (1,())
    \quad
    \text{and}
    \quad
    \funcM{L}^{A}(g)(2,(x,y)) = (2,(x,g(y)))
  \end{equation}
  for all $(x,y) \in A \times B$. Again, see Examples
  \ref{ex:initial-terminal-objects-set}, \ref{ex:product-set}, and
  \ref{ex:coproduct-set} (that is, initial objects, products, and
  coproducts in \set, respectively) if necessary.

\end{example}

\todo{We move on to describe algebra homomorphisms.}

\begin{definition}%%[Algebra homomorphism
  \label{def:algebra-homomorphism}

  %% \parencite[15]{vene-2000}

  Let $\func{F}: \cat{C} \to \cat{C}$ be an endofunctor in a category
  \cat{C}. If $(a,\alpha)$ and $(b,\beta)$ are \func{F}-algebras, an
  \func{F}-algebra homomorphism $f: (a,\alpha) \to (b,\beta)$
  is a morphism $f: a \to b \in \catM{C}$ such that
  \begin{equation}
    \label{eq:algebra-homomorphism}
    \beta \comp \funcM{F}(f) = f \comp \alpha
    \text{,}
  \end{equation}
  that is, such that the diagram in Figure
  \ref{fig:algebra-homomorphism} is commutative. In this case,
  $\dom(f) = (a,\alpha)$ and $\cod(f) = (b,\beta)$.
  \begin{figure}[htbp]
    \begin{center}
      \begin{tikzpicture}
        \node (f a)                {$\funcO{F}(a)$};
        \node (a)   [right of=f a] {$a$};
        \node (f b) [below of=f a] {$\funcO{F}(b)$};
        \node (b)   [right of=f b] {$b$};

        \draw [->] (f a) to node        {$\alpha$} (a);
        \draw [->] (f b) to node [swap] {$\beta$}  (b);

        \draw [->] (a)   to node        {$f$}            (b);
        \draw [->] (f a) to node [swap] {$\funcM{F}(f)$} (f b);
      \end{tikzpicture}
    \end{center}
    \caption{}
    \label{fig:algebra-homomorphism}
  \end{figure}

\end{definition}

\todo{In order to define initial algebras, we need a category of
  algebras and algebra homomorphisms. The following definitions end
  with the definition of such a category.}

\begin{definition}%%[Identity algebra homomorphism]
  \label{def:identity-algebra-homomorphism}

  Let $\func{F}: \cat{C} \to \cat{C}$ be an endofunctor in a category
  \cat{C}. If $(a,\alpha)$ is an \func{F}-algebra, then its identity
  \func{F}-algebra homomorphism $\idO{(a,\alpha)}: (a,\alpha) \to
  (a,\alpha)$ is the identity morphism $\idO{a}: a \to a \in
  \catM{C}$.

  \begin{proof}

    We prove that $\idO{(a,\alpha)}$ is indeed an \func{F}-algebra
    homomorphism:
    \begin{steps}
      \stepm{\alpha \comp \funcM{F}(\idO{a})}
        \eqby{\eqref{eq:functor-identity}}
      \stepm{\alpha \comp \idO{\funcO{F}(a)}}
        \eqby{\eqref{eq:category-identity} with $f = \alpha$}
      \stepm{\idO{a} \comp \alpha}.
    \end{steps}

  \end{proof}

\end{definition}

\begin{definition}%%[Composite algebra homomorphism]
  \label{def:composite-algebra-homomorphism}

  Let $\func{F}: \cat{C} \to \cat{C}$ be an endofunctor in a category
  \cat{C}, and $(a,\alpha)$, $(b,\beta)$, and $(c,\gamma)$ three
  \func{F}-algebras. If $f: (a,\alpha) \to (b,\beta)$ and $g:
  (b,\beta) \to (c,\gamma)$ are two \func{F}-algebra homomorphisms,
  then their composite \func{F}-algebra homomorphism $g \comp f:
  (a,\alpha) \to (c,\gamma)$ is the composite morphism $g \comp f: a
  \to c \in \catM{C}$.

  \begin{proof}

    We prove that $g \comp f$ is indeed an \func{F}-algebra
    homomorphism:
    \begin{steps}
      \stepm{\gamma \comp \funcM{F}(g \comp f)}
        \eqby{\eqref{eq:functor-composition}}
      \stepm{\gamma \comp \funcM{F}(g) \comp \funcM{F}(f)}
        \eqby{\eqref{eq:algebra-homomorphism} with $f = g$}
      \stepm{g \comp \beta \comp \funcM{F}(f)}
        \eqby{\eqref{eq:algebra-homomorphism}}
      \stepm{g \comp f \comp \alpha}.
    \end{steps}

  \end{proof}

\end{definition}

\begin{definition}%% [Category of algebras and algebra homomorphisms
  \label{def:category-alg}

  %% \parencite[16]{vene-2000}

  Let $\func{F}: \cat{C} \to \cat{C}$ be an endofunctor in a category
  \cat{C}. Then \func{F}-\alg is the category of \func{F}-algebras and
  \func{F}-algebra homomorphisms. Its objects are \func{F}-algebras,
  its morphisms are \func{F}-algebra homomorphisms, its identity
  function $\id: \catO{\falg{F}} \to \catM{\falg{F}}$ assigns to each
  \func{F}-algebra $(a, \alpha)$ its identity \func{F}-algebra
  homomorphism, and its composition function $\comp\ : \catM{\falg{F}}
  \times \catM{\falg{F}} \to \catM{\falg{F}}$ assigns to each pair of
  \func{F}-algebra homomorphisms $f$ and $g$ with $\cod(f) = \dom(g)$
  their composite \func{F}-algebra homomorphism.

  \begin{proof}

    Equations \eqref{eq:category-associativity} and
    \eqref{eq:category-identity} prove that \func{F}-\alg is indeed a
    category.

  \end{proof}

\end{definition}

\todo{Introduce the following definition.}

\begin{definition}%% [Initial algebra
  \label{def:initial-algebra}

  %% \parencite[16]{vene-2000}

  Let $\func{F}: \cat{C} \to \cat{C}$ be an endofunctor in a category
  \cat{C}. An \func{F}-algebra $(\mu\func{F},\inmo)$ is the
  initial \func{F}-algebra of the category \func{F}-\alg
  if, for all \func{F}-algebras $(a,\alpha)$, there is a unique
  \func{F}-algebra homomorphism $\cata\alpha: (\mu\func{F},\inmo) \to
  (a,\alpha)$, that is, a morphism $\cata\alpha: \mu\func{F} \to a \in
  \catM{C}$ such that
  \begin{equation}
    \label{eq:cata-charn}
    \alpha \comp \funcM{F}(\cata\alpha) = \cata\alpha \comp \inmo
    \text{,}
  \end{equation}
  or, equivalently, such that the diagram in Figure
  \ref{fig:cata-charn} is commutative. Such an \func{F}-algebra
  homomorphism (that is, a unique \func{F}-algebra homomorphism from
  the initial \func{F}-algebra) is called a catamorphism.
  \begin{figure}[htbp]
    \begin{center}
      \begin{tikzpicture}
        \node (f a)                {$\funcO{F}(\mu\func{F})$};
        \node (a)   [right of=f a] {$\mu\func{F}$};
        \node (f b) [below of=f a] {$\funcO{F}(a)$};
        \node (b)   [right of=f b] {$a$};

        \draw [->] (f a) to node        {$\inmo$} (a);
        \draw [->] (f b) to node [swap] {$\alpha$}  (b);

        \draw [->] (a)   to node        {$\cata\alpha$}            (b);
        \draw [->] (f a) to node [swap] {$\funcM{F}(\cata\alpha)$} (f b);
      \end{tikzpicture}
    \end{center}
    \caption{}
    \label{fig:cata-charn}
  \end{figure}

  Basically, the initial \func{F}-algebra of a category \func{F}-\alg
  is the initial object.

\end{definition}

\todo{Introduce the following lemma.}

\begin{lemma}%% [Reflection and fusion
  \label{lem:cata-refl-fusion}

  %% \parencite[17]{vene-2000}

  Let $\func{F}: \cat{C} \to \cat{C}$ be an endofunctor in a category
  \cat{C}. If $(\mu\func{F},\inmo)$ is the initial \func{F}-algebra of
  the category \func{F}-\alg, then
  \begin{equation}
    \label{eq:cata-refl}
    \idO{\mu\func{F}} = \cata\inmo
  \end{equation}
  and, for all \func{F}-algebra homomorphisms $f: (a,\alpha) \to
  (b,\beta)$,
  \begin{equation}
    \label{eq:cata-fusion}
    f \comp \cata\alpha = \cata\beta
    \text{.}
  \end{equation}

  \begin{proof}

    Since $(\mu\func{F},\inmo)$ is the initial \func{F}-algebra of the
    category \func{F}-\alg, $\cata\inmo$ and $\cata\beta$ are unique.
    As a consequence, \eqref{eq:cata-refl} and \eqref{eq:cata-fusion}
    (see Figure \ref{fig:cata-fusion}) hold.
    \begin{figure}[htbp]
      \begin{center}
        \begin{tikzpicture}
          \node (muf)                  {$\mu\func{F}$};
          \node (f muf) [left of=muf]  {$\funcO{F}(\mu\func{F})$};

          \node (a)     [below of=muf] {$a$};
          \node (f a)   [left of=a]    {$\funcO{F}(a)$};

          \node (b)     [below of=a]   {$b$};
          \node (f b)   [left of=b]    {$\funcO{F}(b)$};

          \node (c)     [right of=a]   {};

          \draw [->] (f muf) to node        {$\inmo$}  (muf);
          \draw [->] (f a)   to node [swap] {$\alpha$} (a);
          \draw [->] (f b)   to node [swap] {$\beta$}  (b);

          \draw [->] (muf) to node {$\cata\alpha$} (a);
          \draw [->] (a)   to node {$f$}           (b);

          \draw [->] (muf) .. controls (c) .. node {$\cata\beta$} (b);

          \draw [->] (f muf) to node [swap] {$\funcM{F}(\cata\alpha)$} (f a);
          \draw [->] (f a)   to node [swap] {$\funcM{F}(f)$}           (f b);
        \end{tikzpicture}
      \end{center}
      \caption{}
      \label{fig:cata-fusion}
    \end{figure}

  \end{proof}

\end{lemma}

Intuitively, the initial algebra denotes the collection of constructor
functions for inductive data types, and the catamorphism is a simple
iteration. The following theorem is ``the formal justification on the
identification of inductive types with initial algebras''
\parencite[17]{vene-2000}.

\begin{theorem}%% [Lambek
  \label{the:lambek}

  %% \parencite[17--18]{vene-2000}

  Let $\func{F}: \cat{C} \to \cat{C}$ be an endofunctor in a category
  \cat{C}. If $(\mu\func{F},\inmo)$ is the initial \func{F}-algebra of
  the category \func{F}-\alg, then $\inmo$ is an isomorphism with its
  inverse
  \begin{equation*}
    \inmo^{-1}: \mu\func{F} \to \funcO{F}(\mu\func{F})
  \end{equation*}
  defined by
  \begin{equation}
    \inmo^{-1} = \cata{\funcM{F}(\inmo)}
    \text{.}
  \end{equation}

  \begin{proof}

    First, we prove that $\inmo \comp \inmo^{-1} = \idO{\mu\func{F}}$,
    that is, $\inmo \comp \cata{\funcM{F}(\inmo)} =
    \idO{\mu\func{F}}$:
    \begin{steps}
      \stepm{\inmo \comp \cata{\funcM{F}(\inmo)}}
        \eqby{\eqref{eq:cata-fusion} with $f = \inmo$}
      \stepm{\cata\inmo}
        \eqby{\eqref{eq:cata-refl}}
      \stepm{\idO{\mu\func{F}}}.
    \end{steps}
    And second, we prove that $\inmo^{-1} \comp \inmo =
    \idO{\funcO{F}(\mu\func{F})}$, that is, $\cata{\funcM{F}(\inmo)}
    \comp \inmo = \idO{\funcO{F}(\mu\func{F})}$:
    \begin{steps}
      \stepm{\cata{\funcM{F}(\inmo)} \comp \inmo}
        \eqby{\eqref{eq:cata-charn} with $\alpha = \funcM{F}(\inmo)$}
      \stepm{\funcM{F}(\inmo) \comp \funcM{F}(\cata{\funcM{F}(\inmo)})}
        \eqby{\eqref{eq:functor-composition} with $f = \cata{\funcM{F}(\inmo)}$ and $g = \inmo$}
      \stepm{\funcM{F}(\inmo \comp \cata{\funcM{F}(\inmo)})}
        \eqtext{see above}
      \stepm{\funcM{F}(\idO{\mu\func{F}})}
        \eqby{\eqref{eq:functor-identity} with $a = \mu\func{F}$}
      \stepm{\idO{\funcO{F}(\mu\func{F})}}.
    \end{steps}

  \end{proof}

\end{theorem}

``The theorem shows that the carrier of the initial algebra is (up to
isomorphism) a fixed point of the functor. In fact, initial algebras
generalize the notion of the least fixed point from lattice theory in
the sense that if the base category is a preorder and thus an
endofunctor is a monotonic function then the carrier of the initial
algebra is the least fixed point of the given function''
\parencite[18]{vene-2000}.

\begin{example}%% [Initial objects
  \label{ex:initial-algebra-initial-object}

  %% \parencite[18]{vene-2000}

  Let \cat{C} be a category with an initial object $0 \in \catO{C}$.
  Then $(0,\idO{0})$, the \func{I}-algebra described in Example
  \ref{ex:algebra-initial-object}, is the initial \func{I}-algebra of
  the category \func{I}-\alg. Indeed, given an \func{I}-algebra
  $(a,\alpha)$, there is a unique \func{I}-algebra homomorphism
  $\cata{\alpha}: (0,\idO{0}) \to (a,\alpha)$ in which the underlying
  morphism $\cata{\alpha}: 0 \to a \in \catM{C}$ is the unique
  morphism given by the fact that $0$ is an initial object. The
  uniqueness of this morphism guarantees that
  \eqref{eq:algebra-homomorphism} holds. In particular, in \set,
  $(\emptyset,\idO{\emptyset})$ is the initial \func{I}-algebra of the
  category \func{I}-\alg.

\end{example}

\begin{example}%% [Natural numbers in $\set$
  \label{ex:initial-algebra-natural-numbers}

  %% \parencite[18--19]{vene-2000}

  In \set, the \func{N}-algebra $(\mathbb{N},[\zero,\suc])$ described
  in Example \ref{ex:algebra-natural-numbers} is the initial
  \func{N}-algebra of the category \func{N}-\alg. For an
  \func{N}-algebra $(A,[z,s])$, that is, a set $A$ and functions $z: 1
  \to A$ and $s: A \to A$, we need a unique \func{N}-algebra
  homomorphism $\cata{[z,s]}$ or
  \begin{equation*}
    \fold(z,s): (\mathbb{N},[\zero,\suc]) \to (A,[z,s])
    \text{,}
  \end{equation*}
  that is, a function $\fold(z,s): \mathbb{N} \to A$ such that
  \begin{equation*}
    \fold(z,s) \comp [\zero,\suc] = [z,s] \comp \funcM{N}(\fold(z,s))
    \text{.}
  \end{equation*}
  Without going into detail, this equation uniquely defines
  $\fold(z,s)$ as
  \begin{equation*}
    \fold(z,s)(\zero()) = z()
  \end{equation*}
  and, for all $n \in \mathbb{N}$,
  \begin{equation*}
    \fold(z,s)(\suc(n)) = s(\fold(z,s)(n))
    \text{,}
  \end{equation*}
  or, more succinctly,
  \begin{equation*}
    \fold(z,s)(n) = s^{n}(z())
    \text{,}
  \end{equation*}
  which yields the required unique \func{N}-algebra homomorphism.

  For instance, addition and multiplication of natural numbers can be
  defined as folds $\add$ and $\mul: \mathbb{N} \times \mathbb{N} \to
  \mathbb{N}$ such that, for all $(m,n) \in \mathbb{N} \times
  \mathbb{N}$,
  \begin{equation*}
    \add(m,n) = \fold(\lambda x.n,\suc)(m)
  \end{equation*}
  and
  \begin{equation*}
    \mul(m,n) = \fold(\zero,\lambda x.\add(m,x))(m)
    \text{,}
  \end{equation*}
  respectively.

\end{example}

\begin{example}%% [Lists in $\set$
  \label{ex:initial-algebra-lists}

  %% \parencite[19--20]{vene-2000}

  In \set, for a set $A$, the $\func{L}^{A}$-algebra
  $(\listof{A},[\nil,\cons])$ described in Example
  \ref{ex:algebra-lists} is the initial algebra of the category
  $\func{L}^{A}$-\alg. Let $(B,[n,c])$ be an algebra over
  $\func{L}^{A}$, that is, a set $A$ and functions $n: 1 \to B$ and
  $c: A \times B \to B$. Then we need a unique $\func{L}^{A}$-algebra
  homomorphism $\fold(n,c): (\listof{A},[\nil,\cons]) \to (B,[n,c])$,
  that is, a function $\foldr(n,c): \listof{A} \to B$ such that
  \begin{equation*}
    \foldr(n,c) \comp [\nil,\cons] = [n,c] \comp \funcM{L}^{A}(\foldr(n,c))
    \text{.}
  \end{equation*}
  This equation uniquely defines $\fold(n,c)$ as
  \begin{equation*}
    \foldr(n,c)(\nil()) = n()
  \end{equation*}
  and, for all $(x,xs) \in A \times \listof{A}$,
  \begin{equation*}
    \foldr(n,c)(\cons(x,xs)) = c(x,\foldr(n,c)(xs))
    \text{,}
  \end{equation*}
  which yields the required unique $\func{L}^{A}$-algebra
  homomorphism.

  As an example, the length of a list of elements of a set $A$ can be
  calculated as a fold $\length: \listof{A} \to \mathbb{N}$ such that
  \begin{equation*}
    \length = \foldr(\zero,\lambda(x,n).\suc(n))
    \text{.}
  \end{equation*}
  As another example, two lists of elements of a set $A$ can be
  appended by a fold $\append: \listof{A} \times \listof{A} \to
  \listof{A}$ such that
  \begin{equation*}
    \append(xs,ys) = \foldr(\lambda x.ys,\cons)(xs)
  \end{equation*}
  for all $(xs,ys) \in \listof{A} \times \listof{A}$. Finally,
  $\map(f): \listof{A} \to \listof{B}$ can be defined as a fold
  \begin{equation*}
    \map(f) = \foldr(\nil,\lambda(x,ys).\cons(f(x),ys))
  \end{equation*}
  for all functions $f: A \to B$.

\end{example}

\section{Algebras in Haskell}
\label{sec:algebras-haskell}

In Haskell, when we define an algebraic data type, its declaration
introduces a new type or type constructor, and zero or more data
constructors. These data specify an initial algebra over an
endofunctor which can be inferred from the information at hand. Given
an algebraic data type, inferring such an endofunctor and proving that
its category of algebras has an initial object amounts to defining a
fold function for that particular type. In fact, such a function is
uniquely determined by the fact that the algebraic data type is an
initial algebra.

As examples, we consider natural numbers and lists. We begin by
describing natural numbers in Haskell in a similar way to that of
Examples \ref{ex:algebra-natural-numbers} and
\ref{ex:initial-algebra-natural-numbers}, that is, natural numbers in
\set.

\begin{example}%% [Natural numbers in Haskell]
  \label{ex:algebra-natural-numbers-haskell}

  In Haskell, natural numbers can be defined as an algebraic data
  type, as follows:
  \begin{codehaskell}
data Nat = Zero | Suc Nat
  \end{codehaskell}
  This declaration introduces a type \texthaskell{Nat} of kind
  \texthaskell{*}, and constructors \texthaskell{Zero} and
  \texthaskell{Suc} with types
  \begin{equation*}
    \text{\texthaskell{Zero :: Nat}}
    \quad
    \text{and}
    \quad
    \text{\texthaskell{Suc :: Nat -> Nat}}
    \text{.}
  \end{equation*}
  These data define an algebra over an endofunctor \texthaskell{N}:
  \begin{codehaskell}
data N a = Z | S a

instance Functor N where
  fmap _ Z     = Z
  fmap f (S x) = S (f x)
\end{codehaskell}
  In detail, this algebra is given by \texthaskell{Nat},
  \texthaskell{Zero}, and \texthaskell{Suc}. Moreover, the algebra in
  question is the initial algebra of the category of algebras over
  \texthaskell{N}:
  \begin{codehaskell}
fold :: a -> (a -> a) -> Nat -> a
fold z s Zero    = z
fold z s (Suc n) = s (fold z s n)
  \end{codehaskell}
  In words, given an algebra over \texthaskell{N}, that is, a type
  \texthaskell{a}, a value \texthaskell{z} of type \texthaskell{a},
  and a function \texthaskell{s} of type \texthaskell{a -> a}, there
  is a unique function of type \texthaskell{Nat -> a}, namely
  \texthaskell{fold z s}.

  For instance, given values \texthaskell{m} and \texthaskell{n} of
  type \texthaskell{Nat}, we define the addition of \texthaskell{m}
  and \texthaskell{n} using \texthaskell{fold n Suc}, that is,
  \texthaskell{fold} for the \texthaskell{N}-algebra specified by
  \texthaskell{Nat}, \texthaskell{n}, and \texthaskell{Suc}, as
  follows:
  \begin{codehaskell}
add :: Nat -> Nat -> Nat
add m n = fold n Suc m
  \end{codehaskell}
  This definition might be easier to understand if we compare it to
  the one yielded by using explicit recursion:
  \begin{codehaskell}
add Zero    n = n
add (Suc m) n = Suc (add m n)
  \end{codehaskell}

  As another example, given values \texthaskell{m} and \texthaskell{n}
  of type \texthaskell{Nat}, we define the multiplication of
  \texthaskell{m} and \texthaskell{n} using \texthaskell{fold Zero
    (add n)}, that is, \texthaskell{fold} for the
  \texthaskell{N}-algebra specified by \texthaskell{Nat},
  \texthaskell{Zero}, and \texthaskell{add n}, as follows:
  \begin{codehaskell}
mul :: Nat -> Nat -> Nat
mul m n = fold Zero (add n) m
  \end{codehaskell}
  In this case, explicit recursion yields the following definition:
  \begin{codehaskell}
mul Zero    n = Zero
mul (Suc m) n = add n (mul m n)
  \end{codehaskell}

\end{example}

Alternatively, we can consider \texthaskell{Nat} and
\begin{equation*}
  \text{\texthaskell{either (\textbackslash () -> Zero) Suc :: Either () Nat -> Nat}}
\end{equation*}
as an algebra over \texthaskell{Either ()} (see Example
\ref{ex:functor-either-haskell}). So, given a type \texthaskell{b} and
\begin{equation*}
  \text{\texthaskell{either (\textbackslash () -> z) s :: Either () a -> a}}
\end{equation*}
we need a unique function \texthaskell{fold} such that
\begin{equation*}
  \text{\texthaskell{fold z s . either (\textbackslash () -> Zero) Suc}}
\end{equation*}
and
\begin{equation*}
  \text{\texthaskell{either (\textbackslash () -> z) s . fmap (fold z s)}}
\end{equation*}
are the same, but that is the above definition of \texthaskell{fold}.

We now describe lists in Haskell in a similar way to that of Examples
\ref{ex:algebra-lists} and \ref{ex:initial-algebra-lists}, that is,
lists in \set.

\begin{example}%% [Lists in Haskell]
  \label{ex:algebra-lists-haskell}

  In Haskell, lists can be defined as an algebraic data type, as
  follows:
  \begin{codehaskell}
data List a = Nil | Cons a (List a)
  \end{codehaskell}
  This declaration introduces a type constructor \texthaskell{List} of
  kind \texthaskell{* -> *}, and constructors \texthaskell{Nil} and
  \texthaskell{Cons} with types
  \begin{equation*}
    \text{\texthaskell{Nil :: List a}}
    \quad
    \text{and}
    \quad
    \text{\texthaskell{Cons :: a -> List a -> List a}}
  \end{equation*}
  for all types \texthaskell{a} of kind \texthaskell{*}. These data
  define an algebra over an endofunctor \texthaskell{L}:
  \begin{codehaskell}
data L a b = N | C a b

instance Functor (L a) where
  fmap _ N       = N
  fmap f (C x y) = C x (f y)
\end{codehaskell}
  More precisely, given a concrete type \texthaskell{a},
  \texthaskell{List a}, \texthaskell{Nil}, and \texthaskell{Cons}
  specify the algebra being discussed. This algebra is the initial
  algebra of the category of algebras over \texthaskell{L}:
  \begin{codehaskell}
foldr :: b -> (a -> b -> b) -> List a -> b
foldr n c Nil         = n
foldr n c (Cons x xs) = c x (foldr n c xs)
  \end{codehaskell}
  That is to say, for a concrete type \texthaskell{a}, given an
  algebra over \texthaskell{L}, that is, a concrete type
  \texthaskell{b}, a value \texthaskell{n} of type \texthaskell{b},
  and a function \texthaskell{c} of type \texthaskell{a -> b -> b},
  there is a unique function of type \texthaskell{List a -> b}, namely
  \texthaskell{foldr n c}.

  As an example, the length of a list of values of a type
  \texthaskell{a} can be calculated using \texthaskell{foldr}, as
  follows:
  \begin{codehaskell}
length :: List a -> Nat
length = foldr Zero (\_ -> Suc)
  \end{codehaskell}
  As another example, two lists \texthaskell{xs} and \texthaskell{ys}
  of values of a type \texthaskell{a} can be appended using
  \texthaskell{fold} for the \texthaskell{L}-algebra specified by
  \texthaskell{List a}, \texthaskell{ys}, and \texthaskell{Cons}:
  \begin{codehaskell}
append :: List a -> List a -> List a
append xs ys = (foldr ys Cons) xs
  \end{codehaskell}
  Finally, given concrete types \texthaskell{a} and \texthaskell{b},
  \texthaskell{map f} can be defined as a \texthaskell{foldr} for all
  functions \texthaskell{f} of type \texthaskell{a -> b}, as follows:
  \begin{codehaskell}
map :: (a -> b) -> List a -> List b
map f = foldr Nil (Cons . f)
  \end{codehaskell}
  Each of these definitions might be easier to understand if we
  compare them to the ones yielded by using explicit recursion:
  \begin{codehaskell}
length Nil         = Zero
length (Cons _ xs) = Suc (length xs)

append Nil         ys = ys
append (Cons x xs) ys = Cons x (concat xs ys)

map _ Nil         = Nil
map f (Cons x xs) = Cons (f x) (fmap f xs)
  \end{codehaskell}

\end{example}

\section{References}
\label{sec:algebras-references}

\todo{...}

\clearemptydoublepage
